---
title: "Palette Colors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Color Palettes
```{r message=FALSE, warning=FALSE}
# 1.Define R Color Data ----
# RGB codes
color.rgb <- t(col2rgb(colors()))

# Hexadecimal codes
color.hex <- rgb(color.rgb[,1], color.rgb[,2], color.rgb[,3], maxColorValue = 255)
# Text highlighting
color.text <- ifelse(apply(color.rgb, 1, mean) > 127, "black", "white")
# Consolidate
color.df <- data.frame(name = colors(),
                       red = color.rgb[, "red"],
                       green = color.rgb[, "green"],
                       blue = color.rgb[, "blue"],
                       hex = color.hex,
                       text = color.text)

# 2.Plot R Colors By Name ----
# configure graphical device
n.col <- 11
n.row <- 60
par(pin = c(11.692, 6.267), mai=c(0.2, 0.2, 0.2, 0.2))
#mai numerical vector indicating margin size c(bottom, left, top, right) in inches  
#pin plot dimensions (width, height) in inches  
#
# create plot
plot(c(0, n.col), c(0, n.row), 
     type = "n", 
     bty = "n", 
     ylab = "", 
     xlab = "", 
     axes = FALSE)
title("R Colors By Name")

for(i in 1:n.col){
     color.count <- (i-1) * n.row
     color.mod <- length(colors()) - color.count
     y.val <- ifelse(color.mod < n.row, n.row - color.mod + 1, 1)
     color.names <- as(color.df[color.count + 1:n.row, "name"], "character")
     rect(i - 1, y.val - 0.5, i, n.row:y.val + 0.5, border = "black", col = color.names)
     text.color <- as(color.df[color.count + 1:n.row, "text"], "character")
     text(i-0.5, n.row:y.val, labels = color.names, cex = 0.5, col = text.color)
}

# 3.Plot R Colors By Hex Code ----
# create plot
plot(c(0, n.col), c(0, n.row), 
     type = "n", 
     bty = "n", 
     ylab = "", 
     xlab = "", 
     axes = FALSE)
title("R Colors By Hex Code")

for(i in 1:n.col){
     color.count <- (i-1) * n.row
     color.mod <- length(colors()) - color.count
     y.val <- ifelse(color.mod < n.row, n.row-color.mod + 1, 1)
     color.names <- as(color.df[color.count + 1:n.row, "hex"], "character")
     rect(i - 1, y.val - 0.5, i, n.row:y.val + 0.5, border = "black", col = color.names)
     text.color <- as(color.df[color.count + 1:n.row, "text"], "character")
     text(i-0.5, n.row:y.val, labels = color.names, cex = 0.5, col = text.color)
}
```

###Alternative Code for Color Palettes

The R function below, SetTextContrastColor, gives a good text color for a given background color name: 
```{r}
SetTextContrastColor <- function(color)
 {
   ifelse( mean(col2rgb(color)) > 127, "black", "white")
 }

# Define this array of text contrast colors that correponds to each
 # member of the colors() array.
 TextContrastColor <- unlist( lapply(colors(), SetTextContrastColor) ) 
```
```{r}
# 1a. Plot matrix of R colors, in index order, 25 per row.
 # This example plots each row of rectangles one at a time.
 colCount <- 25 # number per row
 rowCount <- 27

plot( c(1,colCount), c(0,rowCount), type="n", ylab="", xlab="",
   axes=FALSE, ylim=c(rowCount,0))
 title("R colors") 

for (j in 0:(rowCount-1))
 {
   base <- j*colCount
   remaining <- length(colors()) - base
   RowSize <- ifelse(remaining < colCount, remaining, colCount)
   rect((1:RowSize)-0.5,j-0.5, (1:RowSize)+0.5,j+0.5,
     border="black",
     col=colors()[base + (1:RowSize)])
   text((1:RowSize), j, paste(base + (1:RowSize)), cex=0.7,
     col=TextContrastColor[base + (1:RowSize)])
 } 

```

Alphabetical order is not necessarily a good way to find similar colors.  The RGB values of each of the colors() was converted to hue-saturation-value (HSV) and then sorted by HSV.  This approach groups colors of the same "hue" together a bit better.  Here's the code and graphic produced: 


###Plot matrix of R colors, in "hue" order, 25 per row.
```{r}
 # This example plots each rectangle one at a time.
 RGBColors <- col2rgb(colors()[1:length(colors())])
 HSVColors <- rgb2hsv( RGBColors[1,], RGBColors[2,], RGBColors[3,],
              maxColorValue=255)
 HueOrder <- order( HSVColors[1,], HSVColors[2,], HSVColors[3,] )

plot(0, type="n", ylab="", xlab="",
 axes=FALSE, ylim=c(rowCount,0), xlim=c(1,colCount))

title("R colors -- Sorted by Hue, Saturation, Value")

for (j in 0:(rowCount-1))
 {
   for (i in 1:colCount)
   {
    k <- j*colCount + i
    if (k <= length(colors()))
    {
     rect(i-0.5,j-0.5, i+0.5,j+0.5, border="black", col=colors()[ HueOrder[k] ])
     text(i,j, paste(HueOrder[k]), cex=0.7, col=TextContrastColor[ HueOrder[k] ])
    }
   }
 } 
```
```{r echo=FALSE}
# reset graphical device
dev.off()
```